---
title: "dsImaging validation demo"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{dsImaging validation demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
`%||%` <- function(x, y) if (is.null(x) || !length(x)) y else x
load_imaging_demo_evidence <- function() {
  input_file <- tryCatch(knitr::current_input(dir = TRUE),
                         error = function(e) NULL)
  input_dir <- if (is.null(input_file)) getwd() else dirname(input_file)
  candidates <- c(
    file.path(getwd(), "inst", "extdata",
              "imaging_demo_validation_summary.json"),
    file.path(input_dir, "..", "inst", "extdata",
              "imaging_demo_validation_summary.json"),
    system.file("extdata", "imaging_demo_validation_summary.json",
                package = "dsImagingClient")
  )
  candidates <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(candidates)) {
    stop("Cannot find dsImaging validation evidence JSON.", call. = FALSE)
  }
  jsonlite::fromJSON(candidates[[1]], simplifyVector = FALSE)
}
evidence <- load_imaging_demo_evidence()
stopifnot(isTRUE(evidence$overall_pass))
dim_text <- function(x) paste(unlist(x), collapse = " x ")
site_names <- names(evidence$datasets)
```

This article documents a compact end-to-end validation of the `dsImaging`
runner surface. The demo uses one logical DataSHIELD resource,
`r evidence$resource`, resolved independently by three Opal servers. Each site
holds ten synthetic CT-like NIfTI images, ten matching masks and per-sample
metadata. The validation is deliberately smaller than the LUNG1 cohort article:
its purpose is to exercise the imaging operations provided by `dsImaging`
without making the vignette depend on a large public image download.

The same resource is used for six server-side workflows: dataset validation,
QC metrics, mask connected components, crop-to-mask spatial processing,
deterministic image embeddings, existing-mask radiomics, and threshold
segmentation followed by radiomics. The jobs are submitted from R through
`dsImagingClient`; execution and publication are handled by `dsHPC` inside the
Rock workers.

The evidence bundled with this vignette records `overall_pass =
TRUE`: dataset validation passed on all three servers and
all six workflow families completed with the expected ten local samples per
site.

The three synthetic collections are published with `dsimaging-admin` into the
same S3/MinIO layout as the LUNG1 study, and each Opal resolves the shared
`r evidence$resource` resource to its own local dataset — `imaging_demo_a`,
`imaging_demo_b` and `imaging_demo_c`. The analyst therefore works with a single
resource handle while every site keeps its own images, masks and metadata. The
steps below show what the researcher runs once that configuration is in place.

## Running This Validation

**1. Connect to the DataSHIELD nodes.** Open one session across the three Opal
servers that register the validation resource.

```{r connect, eval=FALSE}
library(DSI)
library(DSOpal)
library(dsImagingClient)
library(dsBaseClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "opal1", url = "https://opal-node-1.example.org",
               user = "researcher", password = "********")
builder$append(server = "opal2", url = "https://opal-node-2.example.org",
               user = "researcher", password = "********")
builder$append(server = "opal3", url = "https://opal-node-3.example.org",
               user = "researcher", password = "********")
conns <- DSI::datashield.login(builder$build(), assign = FALSE)
```

**2. Initialise the resource and resolve each site's dataset.** Bind the shared
resource to a handle, confirm the images and masks validate, and read the
per-site dataset identifier that each Opal resolves the resource to.

```{r init, eval=FALSE}
ds.imaging.init(conns, resource = "dsdemo.imaging_demo", symbol = "img")
ds.imaging.validate(conns, "img")

dataset_ids <- vapply(ds.imaging.metadata(conns, "img"),
                      function(m) m$dataset_id, character(1))
```

**3. Run the six imaging workflows.** Submit each workflow family to every site.
`dsHPC` executes the runners inside the Rock workers and publishes the derived
assets back into each site's catalogue.

```{r workflows, eval=FALSE}
for (srv in names(conns)) {
  cx  <- conns[srv]
  did <- dataset_ids[[srv]]

  ds.imaging.qc.metrics(
    cx, dataset_id = did, image_asset = "images", mask_asset = "masks",
    output_asset = "demo_qc_metrics"
  )
  ds.imaging.mask.operation(
    cx, dataset_id = did, operation = "connected_components",
    mask_asset = "masks", reference_asset = "images",
    min_voxels = 5L, max_components = 1L,
    output_asset = "demo_masks_cc"
  )
  ds.imaging.spatial.process(
    cx, dataset_id = did, image_asset = "images",
    operations = "crop_to_mask", mask_asset = "masks",
    output_asset = "demo_crop_to_mask"
  )
  ds.imaging.embeddings.extract(
    cx, dataset_id = did, image_asset = "images",
    model = "intensity_histogram", bins = 16L,
    output_asset = "demo_embeddings"
  )
  ds.imaging.radiomics.process_collection(
    cx, dataset_id = did,
    segmenter = ds.imaging.segmenter.existing_mask("masks"),
    profile = ds.imaging.radiomics.profile.demo_ct_firstorder()
  )
  ds.imaging.radiomics.segment_and_extract(
    cx, dataset_id = did,
    segmenter = ds.imaging.segmenter.ct_lung_threshold(
      threshold = -320, max_components = 2L, min_voxels = 100L
    ),
    profile = ds.imaging.radiomics.profile.demo_ct_firstorder()
  )
}
```

**4. Load a published asset and inspect it.** Once the jobs finish, load any
published asset server-side and summarise it with ordinary DataSHIELD functions.

```{r inspect, eval=FALSE}
for (srv in names(conns)) {
  ds.imaging.load_asset(conns[srv], dataset_ids[[srv]], "demo_qc_metrics",
                        symbol = "qc", include_metadata = TRUE,
                        syntactic_names = TRUE)
}

ds.dim("qc", datasources = conns)
ds.mean("qc$mask_volume", datasources = conns)

DSI::datashield.logout(conns)
```

## Recorded Results

The values below come from the committed evidence artifact for the run described
above; no live servers are contacted when the article renders.

## Site Layout

```{r site-layout}
site_counts <- data.frame(
  server = site_names,
  dataset = unlist(evidence$datasets, use.names = FALSE),
  samples = as.integer(unlist(evidence$samples_per_site, use.names = FALSE)),
  images = as.integer(unlist(evidence$samples_per_site, use.names = FALSE)),
  masks = as.integer(unlist(evidence$samples_per_site, use.names = FALSE))
)
knitr::kable(site_counts)
```

```{r site-layout-plot, fig.width=7, fig.height=4, fig.alt="Bar chart showing 10 imaging samples on each of three Opal servers."}
if (has_ggplot2) {
  ggplot2::ggplot(site_counts,
                  ggplot2::aes(x = server, y = samples, fill = server)) +
    ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
    ggplot2::geom_text(ggplot2::aes(label = samples), vjust = -0.4) +
    ggplot2::labs(
      x = "DataSHIELD server",
      y = "Synthetic CT images with masks",
      title = "dsImaging validation resource split across three sites"
    ) +
    ggplot2::ylim(0, max(site_counts$samples) * 1.15) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  barplot(site_counts$samples, names.arg = site_counts$server,
          ylab = "Samples", xlab = "DataSHIELD server")
}
```

## Mask QC View

The overlay below is generated from the synthetic validation fixture. It is not
evidence that patient-level overlays should be returned to the analyst; rather,
it shows the kind of local image-mask relation checked inside each site before
only approved derived artefacts and aggregate summaries are exposed.

```{r mask-overlay, fig.width=8.5, fig.alt="Three synthetic CT slices, one per site, with local ROI masks overlaid in red."}
overlay <- "imaging_demo_mask_overlay.png"
if (!file.exists(overlay)) {
  overlay <- system.file("extdata", "imaging_demo_mask_overlay.png",
                         package = "dsImagingClient")
}
knitr::include_graphics(overlay)
```

The same QC pass publishes table assets server-side. The table below reports
site-level aggregate mask summaries computed after loading the QC asset through
DataSHIELD.

```{r mask-aggregates}
mask_aggregates <- data.frame(
  server = site_names,
  mean_mask_voxels = as.numeric(unlist(evidence$qc_aggregates$mask_voxels_mean,
                                       use.names = FALSE)),
  mean_mask_volume = as.numeric(unlist(evidence$qc_aggregates$mask_volume_mean,
                                       use.names = FALSE)),
  mean_masked_hu = as.numeric(unlist(evidence$qc_aggregates$masked_mean_hu,
                                    use.names = FALSE)),
  n_valid = as.integer(unlist(evidence$qc_aggregates$n_valid,
                              use.names = FALSE))
)
knitr::kable(mask_aggregates, digits = 3)
```

```{r mask-aggregates-plot, fig.width=7, fig.height=4, fig.alt="Bar chart of mean local mask volume by DataSHIELD server."}
if (has_ggplot2) {
  ggplot2::ggplot(mask_aggregates,
                  ggplot2::aes(x = server, y = mean_mask_volume,
                               fill = server)) +
    ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
    ggplot2::geom_text(ggplot2::aes(label = round(mean_mask_volume, 1)),
                       vjust = -0.4) +
    ggplot2::labs(
      x = "DataSHIELD server",
      y = "Mean mask volume",
      title = "Mask QC is summarised by site, not by exporting patient images"
    ) +
    ggplot2::ylim(0, max(mask_aggregates$mean_mask_volume) * 1.15) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  barplot(mask_aggregates$mean_mask_volume,
          names.arg = mask_aggregates$server,
          ylab = "Mean mask volume")
}
```

## Workflow Evidence

```{r workflow-table}
workflow_rows <- data.frame(
  workflow = c("QC metrics", "Image embeddings",
               "Existing-mask radiomics", "Segmentation to radiomics"),
  runner_or_route = c("imaging_qc_metrics", "image_embeddings",
                      "existing masks + PyRadiomics",
                      "ct_lung_threshold + PyRadiomics"),
  opal1 = c(dim_text(evidence$dimensions$qc_metrics$opal1),
            dim_text(evidence$dimensions$embeddings$opal1),
            dim_text(evidence$dimensions$existing_mask_radiomics$opal1),
            dim_text(evidence$dimensions$segmented_radiomics$opal1)),
  opal2 = c(dim_text(evidence$dimensions$qc_metrics$opal2),
            dim_text(evidence$dimensions$embeddings$opal2),
            dim_text(evidence$dimensions$existing_mask_radiomics$opal2),
            dim_text(evidence$dimensions$segmented_radiomics$opal2)),
  opal3 = c(dim_text(evidence$dimensions$qc_metrics$opal3),
            dim_text(evidence$dimensions$embeddings$opal3),
            dim_text(evidence$dimensions$existing_mask_radiomics$opal3),
            dim_text(evidence$dimensions$segmented_radiomics$opal3)),
  stringsAsFactors = FALSE
)
knitr::kable(workflow_rows)
```

```{r workflow-status}
status_rows <- do.call(rbind, lapply(names(evidence$workflow_status),
  function(workflow) {
    do.call(rbind, lapply(site_names, function(site) {
      item <- evidence$workflow_status[[workflow]][[site]]
      data.frame(
        workflow = workflow,
        site = site,
        pass = isTRUE(item$pass),
        completed_or_rows = item$completed %||% item$rows,
        failed = item$failed %||% 0,
        stringsAsFactors = FALSE
      )
    }))
  }))
knitr::kable(status_rows)
```

Mask connected-components and crop-to-mask spatial processing publish derived
image or mask assets rather than feature tables. Their job summaries record
ten completed samples and zero failures at each site.

```{r asset-table}
generic <- evidence$generic_assets
asset_row <- function(site, key, label) {
  item <- generic[[site]][[key]]
  summaries <- item$provenance$output$summaries
  summary <- summaries[[names(summaries)[1]]]
  data.frame(
    site = site,
    workflow = label,
    kind = item$kind,
    runner = item$provenance$runner,
    completed = summary$n_done %||% summary$n_images %||% summary$n_samples,
    failed = summary$n_failed %||% 0,
    asset_id = item$asset_id,
    stringsAsFactors = FALSE
  )
}
asset_rows <- do.call(rbind, lapply(site_names, function(site) {
  rbind(
    asset_row(site, "mask_connected_components", "Mask connected components"),
    asset_row(site, "crop_to_mask", "Crop to mask")
  )
}))
knitr::kable(asset_rows[, c("site", "workflow", "kind", "runner",
                            "completed", "failed")])
```

## Radiomics Outputs

The two radiomics passes validate different entry points. The first consumes
the masks already present in the collection. The second creates masks inside
the DataSHIELD job chain using a deterministic CT threshold runner and then
passes those masks to PyRadiomics. Both routes publish ten rows per server.

```{r radiomics-assets}
radiomics_assets <- data.frame(
  site = rep(site_names, times = 2),
  route = rep(c("Existing masks", "Threshold segmentation"), each = length(site_names)),
  asset_id = c(unlist(evidence$assets$existing_mask_radiomics, use.names = FALSE),
               unlist(evidence$assets$segmented_radiomics, use.names = FALSE)),
  dimensions = c(vapply(evidence$dimensions$existing_mask_radiomics, dim_text, character(1)),
                 vapply(evidence$dimensions$segmented_radiomics, dim_text, character(1))),
  stringsAsFactors = FALSE
)
knitr::kable(radiomics_assets)
```

```{r dimensions-plot, fig.width=7, fig.height=4, fig.alt="Bar chart comparing output row counts across QC, embeddings and radiomics workflows."}
plot_dims <- data.frame(
  workflow = rep(workflow_rows$workflow, each = 3),
  server = rep(site_names, times = nrow(workflow_rows)),
  rows = as.integer(c(
    evidence$dimensions$qc_metrics$opal1[[1]],
    evidence$dimensions$qc_metrics$opal2[[1]],
    evidence$dimensions$qc_metrics$opal3[[1]],
    evidence$dimensions$embeddings$opal1[[1]],
    evidence$dimensions$embeddings$opal2[[1]],
    evidence$dimensions$embeddings$opal3[[1]],
    evidence$dimensions$existing_mask_radiomics$opal1[[1]],
    evidence$dimensions$existing_mask_radiomics$opal2[[1]],
    evidence$dimensions$existing_mask_radiomics$opal3[[1]],
    evidence$dimensions$segmented_radiomics$opal1[[1]],
    evidence$dimensions$segmented_radiomics$opal2[[1]],
    evidence$dimensions$segmented_radiomics$opal3[[1]]
  ))
)
if (has_ggplot2) {
  ggplot2::ggplot(plot_dims,
                  ggplot2::aes(x = workflow, y = rows, fill = server)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75),
                      width = 0.68) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      x = NULL,
      y = "Rows published as server-side assets",
      fill = "Server",
      title = "Every validated workflow returns one row per local image"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
} else {
  barplot(matrix(plot_dims$rows, ncol = 3, byrow = TRUE),
          beside = TRUE, names.arg = site_names,
          ylab = "Rows", legend.text = workflow_rows$workflow)
}
```

## Interpretation

The validation exercises the parts of `dsImaging` that matter for downstream
DataSHIELD workflows: a resource-backed image collection can be initialized
from Opal, the worker can materialize MinIO/S3 assets without analyst-side file
access, derived image and mask assets can be published back into the server
catalogue, and tabular outputs can be loaded into the DataSHIELD session for
ordinary aggregate analysis.
