---
title: "Federated LUNG1 radiomics study"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated LUNG1 radiomics study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
load_lung1_evidence <- function() {
  input_file <- tryCatch(knitr::current_input(dir = TRUE),
                         error = function(e) NULL)
  input_dir <- if (is.null(input_file)) getwd() else dirname(input_file)
  candidates <- c(
    system.file("extdata", "lung1_federated_radiomics_results.json",
                package = "dsImagingClient"),
    file.path(getwd(), "inst", "extdata",
              "lung1_federated_radiomics_results.json"),
    file.path(input_dir, "..", "inst", "extdata",
              "lung1_federated_radiomics_results.json")
  )
  candidates <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(candidates)) {
    stop("Cannot find LUNG1 validation evidence JSON.", call. = FALSE)
  }
  jsonlite::fromJSON(candidates[[1]], simplifyDataFrame = TRUE)
}
lung1_evidence <- load_lung1_evidence()
format_num <- function(x, digits = 6) {
  ifelse(abs(x) >= 1e5 | abs(x) < 1e-3,
         formatC(x, format = "e", digits = digits),
         formatC(x, format = "f", digits = digits))
}
feature_label <- function(x) {
  sub("^wavelet\\.HLH_", "wavelet.HLH\n",
      sub("^original_", "original\n", x))
}
```

This article documents the full public LUNG1/NSCLC-Radiomics demonstration for
`dsImagingClient`. The workflow uses CT images and RTSTRUCT `GTV-1` tumour masks
from TCIA, publishes three simulated DataSHIELD sites through
`dsimaging-store`, runs Aerts-signature PyRadiomics extraction through
`dsImaging` and `dsHPC`, and compares the federated summaries with a central
PyRadiomics baseline over the same patients.

The external anchor is the public NSCLC-Radiomics/LUNG1 collection and the
radiomics workflow popularised by Aerts et al. The purpose here is narrower
than re-estimating the complete prognostic modelling programme of that paper:
the validation checks that the image-to-mask-to-radiomics path, executed through
DataSHIELD jobs on separated sites, reproduces the same numerical feature
summaries obtained by a central PyRadiomics run over the same prepared images.

The 422 LUNG1 patients are split reproducibly across three DataSHIELD sites —
Site A holds 142 patients, Site B 143 and Site C 137 — and published with
`dsimaging-admin` into an S3/MinIO-backed layout that is registered as an imaging
resource on each of the three Opal servers. Every site keeps its own CT images,
`GTV-1` tumour masks and sample metadata. The steps below show what the
researcher runs once that server-side configuration is in place; images and masks
are resolved and processed at each site and never reach the client.

## Running This Validation

**1. Connect to the DataSHIELD nodes.** Open one session across the three Opal
servers that register the LUNG1 imaging resource.

```{r connect, eval=FALSE}
library(DSI)
library(DSOpal)
library(dsImagingClient)
library(dsBaseClient)

builder <- DSI::newDSLoginBuilder()
builder$append(server = "opal1", url = "https://opal-node-1.example.org",
               user = "researcher", password = "********")
builder$append(server = "opal2", url = "https://opal-node-2.example.org",
               user = "researcher", password = "********")
builder$append(server = "opal3", url = "https://opal-node-3.example.org",
               user = "researcher", password = "********")
conns <- DSI::datashield.login(builder$build(), assign = FALSE)
```

**2. Initialise the imaging resource.** Bind the LUNG1 resource to a server-side
handle and confirm the CT images and `GTV-1` masks are readable before any job
runs.

```{r init, eval=FALSE}
ds.imaging.init(conns, resource = "dsdemo.lung1_study", symbol = "img")
ds.imaging.validate(conns, "img")
```

**3. Extract the Aerts radiomics signature.** Submit one collection workflow over
the existing tumour masks with the Aerts profile. `dsHPC` schedules the per-image
PyRadiomics jobs at each site; the call returns an opaque workflow symbol.

```{r collection, eval=FALSE}
result <- ds.imaging.radiomics.process_collection(
  conns,
  segmenter  = ds.imaging.segmenter.existing_mask("masks"),
  profile    = ds.imaging.radiomics.profile.aerts_signature(),
  batch_size = 1L,
  timeout    = 0,
  handle     = "img"
)
```

**4. Publish and load the feature table.** Once collection status reports
`is_done = TRUE`, commit the completed outputs to the asset catalogue and load
them — with the joined clinical metadata — into a server-side DataSHIELD table.

```{r load, eval=FALSE}
pub <- stats::setNames(lapply(names(conns), function(server) {
  ds.imaging.radiomics.collection_publish(conns[server], result$symbol)
}), names(conns))

for (server in names(conns)) {
  ds.imaging.radiomics.load_features(
    conns[server],
    asset_id         = pub[[server]]$asset_id,
    symbol           = "rad",
    include_metadata = TRUE,
    syntactic_names  = TRUE,
    handle           = "img"
  )
}
```

**5. Analyse the federated feature table.** `rad` is now an ordinary server-side
DataSHIELD table. Check its dimensions, summarise the Aerts features and fit a
federated two-year-survival GLM from the features and clinical covariates.

```{r analyse, eval=FALSE}
ds.dim("rad", datasources = conns)
ds.mean("rad$original_firstorder_Energy", datasources = conns)

ds.glmSLMA(
  formula = os_2yr_alive ~ original_firstorder_Energy +
    original_shape_Compactness1 +
    original_glrlm_RunLengthNonUniformity +
    wavelet.HLH_glrlm_RunLengthNonUniformity +
    age + gender_male,
  family      = "binomial",
  dataName    = "rad",
  datasources = conns
)

DSI::datashield.logout(conns)
```

## Recorded Results

The values below come from the committed evidence artifact for the run described
above; no live servers are contacted when the article renders.

## Validated Cohort

The final validation pass used every public LUNG1 patient that passed CT and
`GTV-1` mask conversion: `r lung1_evidence$dataset$patients_prepared` patients split across three sites by stable
patient hash.

```{r cohort}
site_counts <- lung1_evidence$site_counts
knitr::kable(site_counts)
```

```{r cohort-plot, fig.width=7, fig.height=4, fig.alt="Bar chart showing the three simulated LUNG1 DataSHIELD sites with 142, 143, and 137 patients."}
if (has_ggplot2) {
  ggplot2::ggplot(site_counts,
                  ggplot2::aes(x = site, y = metadata_rows, fill = site)) +
    ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
    ggplot2::geom_text(ggplot2::aes(label = metadata_rows), vjust = -0.4) +
    ggplot2::labs(
      x = "Simulated site",
      y = "Patients with CT + GTV-1 mask + metadata",
      title = "Full LUNG1 cohort split across three DataSHIELD sites"
    ) +
    ggplot2::ylim(0, max(site_counts$metadata_rows) * 1.12) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  barplot(site_counts$metadata_rows, names.arg = site_counts$site,
          ylab = "Patients", xlab = "Simulated site",
          main = "Full LUNG1 cohort split")
}
```

Published collection assets:

```{r assets}
assets <- lung1_evidence$assets
knitr::kable(assets)
```

After `ds.imaging.radiomics.load_features(..., include_metadata = TRUE)`, the
server-side analysis tables had these dimensions:

```{r dimensions}
loaded_dims <- lung1_evidence$loaded_dimensions
knitr::kable(loaded_dims)
```

## Federated Versus Central Radiomics

The published collection parquet schema is constrained by the Aerts signature
profile to `sample_id` plus four radiomics features. The remaining loaded
columns are clinical metadata joined by `sample_id`.

```{r comparison-data, include=FALSE}
comparison <- lung1_evidence$comparison
comparison$abs_diff_fmt <- format_num(comparison$abs_diff, digits = 3)
comparison$rel_diff_fmt <- format_num(comparison$rel_diff, digits = 3)
comparison$feature_label <- feature_label(comparison$feature)
```

```{r comparison}
knitr::kable(
  comparison[, c("server", "feature", "federated", "central",
                 "abs_diff_fmt", "rel_diff_fmt")],
  col.names = c("server", "feature", "federated mean", "central mean",
                "abs diff", "rel diff"),
  digits = 12
)
```

```{r federated-central-plot, fig.width=8, fig.height=5, fig.alt="Faceted scatter plot comparing central PyRadiomics feature means with federated DataSHIELD means; all points lie on the identity line."}
if (has_ggplot2) {
  ggplot2::ggplot(comparison,
                  ggplot2::aes(x = central, y = federated, color = server)) +
    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                         color = "grey45") +
    ggplot2::geom_point(size = 2.6) +
    ggplot2::facet_wrap(~feature_label, scales = "free") +
    ggplot2::labs(
      x = "Central PyRadiomics mean",
      y = "Federated DataSHIELD mean",
      color = "Server",
      title = "Federated radiomics reproduces the central baseline"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
} else {
  plot(comparison$central, comparison$federated,
       xlab = "Central PyRadiomics mean",
       ylab = "Federated DataSHIELD mean",
       main = "Federated versus central radiomics means")
  abline(0, 1, lty = 2, col = "grey45")
}
```

```{r absolute-error-plot, fig.width=8, fig.height=4.5, fig.alt="Grouped bar chart on a log scale showing very small absolute differences between federated and central feature means."}
plot_errors <- comparison
plot_errors$abs_diff_floor <- pmax(plot_errors$abs_diff, .Machine$double.eps)
if (has_ggplot2) {
  ggplot2::ggplot(plot_errors,
                  ggplot2::aes(x = feature_label, y = abs_diff_floor,
                               fill = server)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75),
                      width = 0.68) +
    ggplot2::scale_y_log10() +
    ggplot2::labs(
      x = NULL,
      y = "Absolute difference, log10 scale",
      fill = "Server",
      title = "Numerical drift stays at floating-point tolerance"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
} else {
  dotchart(plot_errors$abs_diff_floor, labels = plot_errors$feature_label,
           xlab = "Absolute difference",
           main = "Federated-central absolute differences")
}
```

Maximum absolute difference: `r format_num(max(comparison$abs_diff), 3)`.
Maximum relative difference: `r format_num(max(comparison$rel_diff), 3)`.

## Clinical Metadata And Federated GLM

Clinical metadata were joined to the radiomics feature table on the server side
before analysis.

```{r clinical}
clinical <- lung1_evidence$clinical
knitr::kable(clinical, digits = 7)
```

```{r clinical-plot, fig.width=7, fig.height=4, fig.alt="Faceted bar chart of site-level means for survival time and two-year overall survival status."}
clinical_long <- data.frame(
  metric = rep(clinical$metric, each = 3),
  server = rep(c("opal1", "opal2", "opal3"), times = nrow(clinical)),
  value = as.numeric(t(clinical[, c("opal1", "opal2", "opal3")]))
)
if (has_ggplot2) {
  ggplot2::ggplot(clinical_long,
                  ggplot2::aes(x = server, y = value, fill = server)) +
    ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
    ggplot2::facet_wrap(~metric, scales = "free_y") +
    ggplot2::labs(
      x = "Server",
      y = "Federated site mean",
      title = "Clinical metadata remain available after server-side joining"
    ) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  barplot(as.matrix(clinical[, c("opal1", "opal2", "opal3")]),
          beside = TRUE, legend.text = clinical$metric,
          ylab = "Site mean", main = "Clinical metadata site means")
}
```

A federated `ds.glmSLMA()` model was fit with:

```{r glm-formula}
stats::as.formula(lung1_evidence$glm$formula)
```

The run produced `r paste0("num.valid.studies = ", lung1_evidence$glm$num_valid_studies)`;
`<new.glm.obj>` was created and validated in all data sources.

```{r glm-results}
glm_fe <- lung1_evidence$glm$fixed_effects
knitr::kable(glm_fe, digits = 6)
```

```{r glm-z-plot, fig.width=7, fig.height=4.5, fig.alt="Horizontal bar chart of fixed-effect coefficient divided by standard error for each federated GLM term, with dashed reference lines at plus or minus 1.96."}
glm_plot <- glm_fe
glm_plot$z_score <- glm_plot$pooled.FE / glm_plot$se.FE
glm_plot$term <- factor(glm_plot$term,
                        levels = glm_plot$term[order(glm_plot$z_score)])
if (has_ggplot2) {
  ggplot2::ggplot(glm_plot, ggplot2::aes(x = term, y = z_score)) +
    ggplot2::geom_hline(yintercept = c(-1.96, 1.96), linetype = "dashed",
                        color = "grey45") +
    ggplot2::geom_col(fill = "#3b6ea8", width = 0.7) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      x = NULL,
      y = "Fixed-effect coefficient / standard error",
      title = "Federated GLM coefficient signal"
    ) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  dotchart(glm_plot$z_score, labels = as.character(glm_plot$term),
           xlab = "Coefficient / standard error",
           main = "Federated GLM coefficient signal")
  abline(v = c(-1.96, 1.96), lty = 2, col = "grey45")
}
```

## Operational Notes

- `process_collection()` exposes only its coarse state and completion bit; it
  does not return per-site sample, patient, or job counts.
- Keep the opaque workflow symbol returned by each call. Cross-workflow job
  listing is deliberately unavailable to analysts.
