---
title: "Federated LUNG1 radiomics study"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated LUNG1 radiomics study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
format_num <- function(x, digits = 6) {
  ifelse(abs(x) >= 1e5 | abs(x) < 1e-3,
         formatC(x, format = "e", digits = digits),
         formatC(x, format = "f", digits = digits))
}
```

This article documents the full public LUNG1/NSCLC-Radiomics demonstration for
`dsImagingClient`. The workflow uses CT images and RTSTRUCT `GTV-1` tumour masks
from TCIA, publishes three simulated DataSHIELD sites through
`dsimaging-store`, runs Aerts-signature PyRadiomics extraction through
`dsImaging` and `dsHPC`, and compares the federated summaries with a central
PyRadiomics baseline over the same patients.

The default pkgdown build renders the validated results below without launching
jobs. To execute the live demo while rendering this article, set
`DSIMAGINGCLIENT_RUN_LUNG1_VIGNETTE=true` and point `LUNG1_WORKDIR` at a
prepared LUNG1 workspace.

## Running The Demo

The preparation and execution scripts are installed with the package under
`inst/demos/lung1_federated_study`.

```{r demo-paths}
demo_dir <- system.file("demos", "lung1_federated_study",
                        package = "dsImagingClient")
if (!nzchar(demo_dir)) {
  demo_dir <- normalizePath(file.path("..", "inst", "demos",
                                     "lung1_federated_study"),
                            mustWork = FALSE)
}
prepare_script <- file.path(demo_dir, "prepare_lung1_study.py")
run_script <- file.path(demo_dir, "run_lung1_datashield.R")
c(prepare_script = prepare_script, run_script = run_script)
```

Prepare the full public cohort:

```{bash prepare-full, eval=FALSE}
python3 inst/demos/lung1_federated_study/prepare_lung1_study.py \
  --workdir /tmp/dsimaging_lung1_full \
  --all-patients
```

Run or resume the federated DataSHIELD pass:

```{bash run-demo, eval=FALSE}
LUNG1_WORKDIR=/tmp/dsimaging_lung1_full \
LUNG1_DATASET_PREFIX=lung1_full_site \
LUNG1_OPAL_RESOURCE=lung1_full_study \
LUNG1_PUBLISH=FALSE \
LUNG1_RUN_JOBS=FALSE \
LUNG1_TIMEOUT=0 \
LUNG1_RUN_GLM=TRUE \
OPAL_USER=administrator \
OPAL_PASSWORD=admin123 \
Rscript inst/demos/lung1_federated_study/run_lung1_datashield.R
```

```{r run-live-demo, eval=identical(tolower(Sys.getenv("DSIMAGINGCLIENT_RUN_LUNG1_VIGNETTE")), "true")}
demo_env <- c(
  paste0("LUNG1_WORKDIR=", Sys.getenv("LUNG1_WORKDIR",
                                      "/tmp/dsimaging_lung1_full")),
  paste0("LUNG1_DATASET_PREFIX=", Sys.getenv("LUNG1_DATASET_PREFIX",
                                             "lung1_full_site")),
  paste0("LUNG1_OPAL_RESOURCE=", Sys.getenv("LUNG1_OPAL_RESOURCE",
                                            "lung1_full_study")),
  paste0("LUNG1_PUBLISH=", Sys.getenv("LUNG1_PUBLISH", "FALSE")),
  paste0("LUNG1_RUN_JOBS=", Sys.getenv("LUNG1_RUN_JOBS", "FALSE")),
  paste0("LUNG1_TIMEOUT=", Sys.getenv("LUNG1_TIMEOUT", "0")),
  paste0("LUNG1_RUN_GLM=", Sys.getenv("LUNG1_RUN_GLM", "TRUE")),
  paste0("OPAL_USER=", Sys.getenv("OPAL_USER", "administrator")),
  paste0("OPAL_PASSWORD=", Sys.getenv("OPAL_PASSWORD", "admin123"))
)
status <- system2("Rscript", run_script, env = demo_env)
if (!identical(status, 0L)) {
  stop("LUNG1 demo failed with exit status ", status, call. = FALSE)
}
```

## Validated Cohort

The final validation pass used every public LUNG1 patient that passed CT and
`GTV-1` mask conversion: `422` patients split across three sites by stable
patient hash.

```{r cohort}
site_counts <- data.frame(
  site = c("site_a", "site_b", "site_c"),
  dataset = c("lung1_full_site_a", "lung1_full_site_b", "lung1_full_site_c"),
  images = c(142L, 143L, 137L),
  masks = c(142L, 143L, 137L),
  metadata_rows = c(142L, 143L, 137L)
)
knitr::kable(site_counts)
```

Published collection assets:

```{r assets}
assets <- data.frame(
  opal = c("opal1", "opal2", "opal3"),
  dataset = c("lung1_full_site_a", "lung1_full_site_b", "lung1_full_site_c"),
  generation = c("gen_20260509_152105_9e354de9",
                 "gen_20260509_152108_8cf5c55f",
                 "gen_20260509_152110_b82b5785"),
  asset = c("asset_20260509_165902_42b9b1a5",
            "asset_20260509_170056_0bfdbf8e",
            "asset_20260509_170057_9a3db2ab")
)
knitr::kable(assets)
```

After `ds.imaging.radiomics.load_features(..., include_metadata = TRUE)`, the
server-side analysis tables had these dimensions:

```{r dimensions}
loaded_dims <- data.frame(
  source = c("opal1", "opal2", "opal3", "combined"),
  dimensions = c("142 x 20", "143 x 20", "137 x 20", "422 x 20")
)
knitr::kable(loaded_dims)
```

## Federated Versus Central Radiomics

The published collection parquet schema is constrained by the Aerts signature
profile to `sample_id` plus four radiomics features. The remaining loaded
columns are clinical metadata joined by `sample_id`.

```{r comparison-data, include=FALSE}
embedded_comparison <- data.frame(
  server = c("opal1", "opal1", "opal1", "opal1",
             "opal2", "opal2", "opal2", "opal2",
             "opal3", "opal3", "opal3", "opal3"),
  feature = c(
    "original_firstorder_Energy",
    "original_glrlm_RunLengthNonUniformity",
    "original_shape_Compactness1",
    "wavelet.HLH_glrlm_RunLengthNonUniformity",
    "original_firstorder_Energy",
    "original_glrlm_RunLengthNonUniformity",
    "original_shape_Compactness1",
    "wavelet.HLH_glrlm_RunLengthNonUniformity",
    "original_firstorder_Energy",
    "original_glrlm_RunLengthNonUniformity",
    "original_shape_Compactness1",
    "wavelet.HLH_glrlm_RunLengthNonUniformity"
  ),
  federated = c(
    912743691.507042, 12430.8791801311, 0.0258847932464428,
    9307.55647427235, 924557312.867133, 13151.8800546352,
    0.0268293770986911, 10245.5239331547, 1186055557.30657,
    14466.9012668664, 0.0263166107584965, 11379.0070861031
  ),
  central = c(
    912743691.507043, 12430.8791801311, 0.0258847932464428,
    9307.55647427235, 924557312.867133, 13151.8800546352,
    0.0268293770986911, 10245.5239331547, 1186055557.30657,
    14466.9012668664, 0.0263166107584965, 11379.0070861031
  ),
  abs_diff = c(
    5.96046447753906e-07, 4.36557456851006e-11, 2.42861286636753e-17,
    1.81898940354586e-12, 1.19209289550781e-07, 1.63709046319127e-11,
    1.38777878078145e-17, 4.91127138957381e-11, 9.5367431640625e-07,
    1.81898940354586e-11, 4.85722573273506e-17, 7.27595761418343e-12
  ),
  rel_diff = c(
    6.53027189670045e-16, 3.51187917222121e-15, 9.38239236931621e-16,
    1.95431465667047e-16, 1.28936614195504e-16, 1.24475775051971e-15,
    5.17260902359582e-16, 4.79357758726308e-15, 8.04072212748586e-16,
    1.25734555727694e-15, 1.84568817668395e-15, 6.39419376324089e-16
  )
)

workdir <- Sys.getenv("LUNG1_WORKDIR", "/tmp/dsimaging_lung1_full")
comparison_path <- file.path(workdir, "federated_vs_central_means.csv")
comparison <- if (file.exists(comparison_path)) {
  read.csv(comparison_path, check.names = FALSE)
} else {
  embedded_comparison
}
comparison$abs_diff_fmt <- format_num(comparison$abs_diff, digits = 3)
comparison$rel_diff_fmt <- format_num(comparison$rel_diff, digits = 3)
```

```{r comparison}
knitr::kable(
  comparison[, c("server", "feature", "federated", "central",
                 "abs_diff_fmt", "rel_diff_fmt")],
  col.names = c("server", "feature", "federated mean", "central mean",
                "abs diff", "rel diff"),
  digits = 12
)
```

Maximum absolute difference: `r format_num(max(comparison$abs_diff), 3)`.
Maximum relative difference: `r format_num(max(comparison$rel_diff), 3)`.

## Clinical Metadata And Federated GLM

Clinical metadata were joined to the radiomics feature table on the server side
before analysis.

```{r clinical}
clinical <- data.frame(
  metric = c("mean survival_time_days", "mean os_2yr_alive"),
  opal1 = c(1000.6197, 0.4184397),
  opal2 = c(976.9650, 0.4055944),
  opal3 = c(989.0803, 0.3823529)
)
knitr::kable(clinical, digits = 7)
```

A federated `ds.glmSLMA()` model was fit with:

```{r glm-formula}
glm_formula <- os_2yr_alive ~ original_firstorder_Energy +
  original_shape_Compactness1 +
  original_glrlm_RunLengthNonUniformity +
  wavelet.HLH_glrlm_RunLengthNonUniformity +
  age + gender_male
glm_formula
```

The run produced `num.valid.studies = 3`; `<new.glm.obj>` was created and
validated in all data sources.

```{r glm-results}
glm_fe <- data.frame(
  term = c("(Intercept)", "original_firstorder_Energy",
           "original_shape_Compactness1",
           "original_glrlm_RunLengthNonUniformity",
           "wavelet.HLH_glrlm_RunLengthNonUniformity",
           "age", "gender_male"),
  pooled.FE = c(1.416178e-01, -8.452579e-11, 2.383309e+01,
                8.191498e-05, -1.184013e-04, -9.693862e-03,
                -1.464250e-01),
  se.FE = c(9.263942e-01, 1.228317e-10, 2.154354e+01,
            5.991988e-05, 7.716003e-05, 1.150634e-02,
            2.446642e-01)
)
knitr::kable(glm_fe, digits = 6)
```

## Operational Notes

- The full run is resumable. DICOM downloads, NIfTI/mask conversion, central
  PyRadiomics extraction, DataSHIELD job submission, and collection publishing
  can be resumed from the same `LUNG1_WORKDIR`.
- `process_collection()` reports disclosure-bucketed metadata, so its status may
  show `128` rather than exact per-site counts. `ds.dim("rad")` and the local
  manifest are the engineering checks for exact dimensions.
- Admin job listing and cancellation use `dshpc.admin_key` or
  `DSHPC_ADMIN_KEY`; wrong keys are rejected by all sites.
