---
title: "Federated LUNG1 radiomics study"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Federated LUNG1 radiomics study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
load_lung1_evidence <- function() {
  input_file <- tryCatch(knitr::current_input(dir = TRUE),
                         error = function(e) NULL)
  input_dir <- if (is.null(input_file)) getwd() else dirname(input_file)
  candidates <- c(
    system.file("extdata", "lung1_federated_radiomics_results.json",
                package = "dsImagingClient"),
    file.path(getwd(), "inst", "extdata",
              "lung1_federated_radiomics_results.json"),
    file.path(input_dir, "..", "inst", "extdata",
              "lung1_federated_radiomics_results.json")
  )
  candidates <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(candidates)) {
    stop("Cannot find LUNG1 validation evidence JSON.", call. = FALSE)
  }
  jsonlite::fromJSON(candidates[[1]], simplifyDataFrame = TRUE)
}
lung1_evidence <- load_lung1_evidence()
format_num <- function(x, digits = 6) {
  ifelse(abs(x) >= 1e5 | abs(x) < 1e-3,
         formatC(x, format = "e", digits = digits),
         formatC(x, format = "f", digits = digits))
}
feature_label <- function(x) {
  sub("^wavelet\\.HLH_", "wavelet.HLH\n",
      sub("^original_", "original\n", x))
}
```

This article documents the full public LUNG1/NSCLC-Radiomics demonstration for
`dsImagingClient`. The workflow uses CT images and RTSTRUCT `GTV-1` tumour masks
from TCIA, publishes three simulated DataSHIELD sites through
`dsimaging-store`, runs Aerts-signature PyRadiomics extraction through
`dsImaging` and `dsHPC`, and compares the federated summaries with a central
PyRadiomics baseline over the same patients.

The external anchor is the public NSCLC-Radiomics/LUNG1 collection and the
radiomics workflow popularised by Aerts et al. The purpose here is narrower
than re-estimating the complete prognostic modelling programme of that paper:
the validation checks that the image-to-mask-to-radiomics path, executed through
DataSHIELD jobs on separated sites, reproduces the same numerical feature
summaries obtained by a central PyRadiomics run over the same prepared images.

The default pkgdown build renders the validated results below without launching
jobs. To execute the live demo while rendering this article, set
`DSIMAGINGCLIENT_RUN_LUNG1_VIGNETTE=true` and point `LUNG1_WORKDIR` at a
prepared LUNG1 workspace.

## Running The Demo

The preparation and execution scripts are installed with the package under
`inst/demos/lung1_federated_study`.

```{r demo-paths}
demo_dir <- system.file("demos", "lung1_federated_study",
                        package = "dsImagingClient")
if (!nzchar(demo_dir)) {
  demo_dir <- normalizePath(file.path("..", "inst", "demos",
                                     "lung1_federated_study"),
                            mustWork = FALSE)
}
prepare_script <- file.path(demo_dir, "prepare_lung1_study.py")
run_script <- file.path(demo_dir, "run_lung1_datashield.R")
c(prepare_script = prepare_script, run_script = run_script)
```

Prepare the full public cohort:

```{bash prepare-full, eval=FALSE}
python3 inst/demos/lung1_federated_study/prepare_lung1_study.py \
  --workdir /tmp/dsimaging_lung1_full \
  --all-patients
```

For a metadata-only check before downloading DICOM files:

```{bash prepare-plan, eval=FALSE}
python3 inst/demos/lung1_federated_study/prepare_lung1_study.py \
  --workdir /tmp/dsimaging_lung1_plan \
  --all-patients \
  --dry-run
```

Run or resume the federated DataSHIELD pass:

```{bash run-demo, eval=FALSE}
LUNG1_WORKDIR=/tmp/dsimaging_lung1_full \
LUNG1_DATASET_PREFIX=lung1_full_site \
LUNG1_OPAL_RESOURCE=lung1_full_study \
LUNG1_PUBLISH=FALSE \
LUNG1_RUN_JOBS=FALSE \
LUNG1_TIMEOUT=0 \
LUNG1_RUN_GLM=TRUE \
OPAL_USER=administrator \
OPAL_PASSWORD=admin123 \
Rscript inst/demos/lung1_federated_study/run_lung1_datashield.R
```

```{r run-live-demo, eval=identical(tolower(Sys.getenv("DSIMAGINGCLIENT_RUN_LUNG1_VIGNETTE")), "true")}
demo_env <- c(
  paste0("LUNG1_WORKDIR=", Sys.getenv("LUNG1_WORKDIR",
                                      "/tmp/dsimaging_lung1_full")),
  paste0("LUNG1_DATASET_PREFIX=", Sys.getenv("LUNG1_DATASET_PREFIX",
                                             "lung1_full_site")),
  paste0("LUNG1_OPAL_RESOURCE=", Sys.getenv("LUNG1_OPAL_RESOURCE",
                                            "lung1_full_study")),
  paste0("LUNG1_PUBLISH=", Sys.getenv("LUNG1_PUBLISH", "FALSE")),
  paste0("LUNG1_RUN_JOBS=", Sys.getenv("LUNG1_RUN_JOBS", "FALSE")),
  paste0("LUNG1_TIMEOUT=", Sys.getenv("LUNG1_TIMEOUT", "0")),
  paste0("LUNG1_RUN_GLM=", Sys.getenv("LUNG1_RUN_GLM", "TRUE")),
  paste0("OPAL_USER=", Sys.getenv("OPAL_USER", "administrator")),
  paste0("OPAL_PASSWORD=", Sys.getenv("OPAL_PASSWORD", "admin123"))
)
status <- system2("Rscript", run_script, env = demo_env)
if (!identical(status, 0L)) {
  stop("LUNG1 demo failed with exit status ", status, call. = FALSE)
}
```

## Validated Cohort

The final validation pass used every public LUNG1 patient that passed CT and
`GTV-1` mask conversion: `r lung1_evidence$dataset$patients_prepared` patients split across three sites by stable
patient hash.

```{r cohort}
site_counts <- lung1_evidence$site_counts
knitr::kable(site_counts)
```

```{r cohort-plot, fig.width=7, fig.height=4, fig.alt="Bar chart showing the three simulated LUNG1 DataSHIELD sites with 142, 143, and 137 patients."}
if (has_ggplot2) {
  ggplot2::ggplot(site_counts,
                  ggplot2::aes(x = site, y = metadata_rows, fill = site)) +
    ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
    ggplot2::geom_text(ggplot2::aes(label = metadata_rows), vjust = -0.4) +
    ggplot2::labs(
      x = "Simulated site",
      y = "Patients with CT + GTV-1 mask + metadata",
      title = "Full LUNG1 cohort split across three DataSHIELD sites"
    ) +
    ggplot2::ylim(0, max(site_counts$metadata_rows) * 1.12) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  barplot(site_counts$metadata_rows, names.arg = site_counts$site,
          ylab = "Patients", xlab = "Simulated site",
          main = "Full LUNG1 cohort split")
}
```

Published collection assets:

```{r assets}
assets <- lung1_evidence$assets
knitr::kable(assets)
```

After `ds.imaging.radiomics.load_features(..., include_metadata = TRUE)`, the
server-side analysis tables had these dimensions:

```{r dimensions}
loaded_dims <- lung1_evidence$loaded_dimensions
knitr::kable(loaded_dims)
```

## Federated Versus Central Radiomics

The published collection parquet schema is constrained by the Aerts signature
profile to `sample_id` plus four radiomics features. The remaining loaded
columns are clinical metadata joined by `sample_id`.

```{r comparison-data, include=FALSE}
workdir <- Sys.getenv("LUNG1_WORKDIR", "/tmp/dsimaging_lung1_full")
comparison_path <- file.path(workdir, "federated_vs_central_means.csv")
comparison <- if (file.exists(comparison_path)) {
  read.csv(comparison_path, check.names = FALSE)
} else {
  lung1_evidence$comparison
}
comparison$abs_diff_fmt <- format_num(comparison$abs_diff, digits = 3)
comparison$rel_diff_fmt <- format_num(comparison$rel_diff, digits = 3)
comparison$feature_label <- feature_label(comparison$feature)
```

```{r comparison}
knitr::kable(
  comparison[, c("server", "feature", "federated", "central",
                 "abs_diff_fmt", "rel_diff_fmt")],
  col.names = c("server", "feature", "federated mean", "central mean",
                "abs diff", "rel diff"),
  digits = 12
)
```

```{r federated-central-plot, fig.width=8, fig.height=5, fig.alt="Faceted scatter plot comparing central PyRadiomics feature means with federated DataSHIELD means; all points lie on the identity line."}
if (has_ggplot2) {
  ggplot2::ggplot(comparison,
                  ggplot2::aes(x = central, y = federated, color = server)) +
    ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                         color = "grey45") +
    ggplot2::geom_point(size = 2.6) +
    ggplot2::facet_wrap(~feature_label, scales = "free") +
    ggplot2::labs(
      x = "Central PyRadiomics mean",
      y = "Federated DataSHIELD mean",
      color = "Server",
      title = "Federated radiomics reproduces the central baseline"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
} else {
  plot(comparison$central, comparison$federated,
       xlab = "Central PyRadiomics mean",
       ylab = "Federated DataSHIELD mean",
       main = "Federated versus central radiomics means")
  abline(0, 1, lty = 2, col = "grey45")
}
```

```{r absolute-error-plot, fig.width=8, fig.height=4.5, fig.alt="Grouped bar chart on a log scale showing very small absolute differences between federated and central feature means."}
plot_errors <- comparison
plot_errors$abs_diff_floor <- pmax(plot_errors$abs_diff, .Machine$double.eps)
if (has_ggplot2) {
  ggplot2::ggplot(plot_errors,
                  ggplot2::aes(x = feature_label, y = abs_diff_floor,
                               fill = server)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.75),
                      width = 0.68) +
    ggplot2::scale_y_log10() +
    ggplot2::labs(
      x = NULL,
      y = "Absolute difference, log10 scale",
      fill = "Server",
      title = "Numerical drift stays at floating-point tolerance"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(legend.position = "bottom")
} else {
  dotchart(plot_errors$abs_diff_floor, labels = plot_errors$feature_label,
           xlab = "Absolute difference",
           main = "Federated-central absolute differences")
}
```

Maximum absolute difference: `r format_num(max(comparison$abs_diff), 3)`.
Maximum relative difference: `r format_num(max(comparison$rel_diff), 3)`.

## Clinical Metadata And Federated GLM

Clinical metadata were joined to the radiomics feature table on the server side
before analysis.

```{r clinical}
clinical <- lung1_evidence$clinical
knitr::kable(clinical, digits = 7)
```

```{r clinical-plot, fig.width=7, fig.height=4, fig.alt="Faceted bar chart of site-level means for survival time and two-year overall survival status."}
clinical_long <- data.frame(
  metric = rep(clinical$metric, each = 3),
  server = rep(c("opal1", "opal2", "opal3"), times = nrow(clinical)),
  value = as.numeric(t(clinical[, c("opal1", "opal2", "opal3")]))
)
if (has_ggplot2) {
  ggplot2::ggplot(clinical_long,
                  ggplot2::aes(x = server, y = value, fill = server)) +
    ggplot2::geom_col(width = 0.65, show.legend = FALSE) +
    ggplot2::facet_wrap(~metric, scales = "free_y") +
    ggplot2::labs(
      x = "Server",
      y = "Federated site mean",
      title = "Clinical metadata remain available after server-side joining"
    ) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  barplot(as.matrix(clinical[, c("opal1", "opal2", "opal3")]),
          beside = TRUE, legend.text = clinical$metric,
          ylab = "Site mean", main = "Clinical metadata site means")
}
```

A federated `ds.glmSLMA()` model was fit with:

```{r glm-formula}
stats::as.formula(lung1_evidence$glm$formula)
```

The run produced `r paste0("num.valid.studies = ", lung1_evidence$glm$num_valid_studies)`;
`<new.glm.obj>` was created and validated in all data sources.

```{r glm-results}
glm_fe <- lung1_evidence$glm$fixed_effects
knitr::kable(glm_fe, digits = 6)
```

```{r glm-z-plot, fig.width=7, fig.height=4.5, fig.alt="Horizontal bar chart of fixed-effect coefficient divided by standard error for each federated GLM term, with dashed reference lines at plus or minus 1.96."}
glm_plot <- glm_fe
glm_plot$z_score <- glm_plot$pooled.FE / glm_plot$se.FE
glm_plot$term <- factor(glm_plot$term,
                        levels = glm_plot$term[order(glm_plot$z_score)])
if (has_ggplot2) {
  ggplot2::ggplot(glm_plot, ggplot2::aes(x = term, y = z_score)) +
    ggplot2::geom_hline(yintercept = c(-1.96, 1.96), linetype = "dashed",
                        color = "grey45") +
    ggplot2::geom_col(fill = "#3b6ea8", width = 0.7) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      x = NULL,
      y = "Fixed-effect coefficient / standard error",
      title = "Federated GLM coefficient signal"
    ) +
    ggplot2::theme_minimal(base_size = 12)
} else {
  dotchart(glm_plot$z_score, labels = as.character(glm_plot$term),
           xlab = "Coefficient / standard error",
           main = "Federated GLM coefficient signal")
  abline(v = c(-1.96, 1.96), lty = 2, col = "grey45")
}
```

## Operational Notes

- The full run is resumable. DICOM downloads, NIfTI/mask conversion, central
  PyRadiomics extraction, DataSHIELD job submission, and collection publishing
  can be resumed from the same `LUNG1_WORKDIR`.
- `process_collection()` reports disclosure-bucketed metadata, so its status may
  show `128` rather than exact per-site counts. `ds.dim("rad")` and the local
  manifest are the engineering checks for exact dimensions.
- Admin job listing and cancellation use `dshpc.admin_key` or
  `DSHPC_ADMIN_KEY`; wrong keys are rejected by all sites.
